OPTIMAL PRICING AND INFORMAL SHARING: \\ EVIDENCE FROM PIPED WATER IN MANILA

### 2018-12-13 phil_paper/jmp_wjv.tex
jmp fixes for qje reject

	In developing cities, high costs of connecting to piped water lead many households to use low-quality, non-piped sources.  In response, cities often set low fixed prices so that more households connect and high marginal prices to cover production costs.  This paper identifies a common alternative source --- sharing a neighbor's connection --- that has large implications for these pricing strategies.  With microdata from 1.5 million water connections in Manila, I structurally estimate water demand in the presence of water sharing and solve the regulator's problem: setting prices to maximize consumer surplus while covering production costs.  I find that low fixed and high marginal prices weaken incentives to share water between neighbors.  In contrast, the optimal pricing policy features high fixed and low marginal prices, increases shared connections, and ensures nearly universal access to piped water compared to current prices in Manila.

@@

These estimates allow me to solve a standard constrained regulator pricing problem: setting prices to cover fixed and marginal costs while maximizing consumer surplus.  I also let the regulator place extra weight on surplus for smaller, poorer users.  In the presence of sharing, the standard approach --- low fixed fees and high marginal prices --- excludes many small-users from piped water by reducing incentives to share connections.  Instead, these households substitute toward non-piped water sources, which are easy to access, have relatively lower marginal prices, and provide varying water quality.\footnote{Common non-piped water sources in Manila include deepwells, tanker trucks, and water refilling stations.  Without fixed prices, land-tenure requirements, or administrative expenses, these sources often have low costs of access.}  The opposite approach --- high fixed fees and low marginal prices --- maximizes both access to piped water and welfare even for smaller users.  High fixed fees are paid in full by households who connect individually and are split between sharing households.  Informal sharing arrangements enable small-users to access piped water from their neighbors for relatively lower marginal prices than those offered by non-piped sources.  By reducing the total number of connections, sharing arrangements also let the city save on installation and maintenance costs, providing an efficient means for extending piped water to low-demand households.\footnote{In this regulatory arrangement, cost savings are passed on to consumers through lower prices.}

@@

To estimate the marginal hassle costs of sharing water, the second step analyzes a quasi-experiment where breaks in service pipes force some households to switch from their own water connections to their neighbors' connections.  I then measure how much less water households use from their neighbors' connections and attribute this gap to the marginal hassle costs of sharing.  This exercise recovers sizable marginal hassle costs on the order of \input{tables/hassle_percentage}\unskip\% of the average marginal water price.\footnote{This revealed-preference approach builds on a descriptive literature that identifies potential sources of water sharing frictions (\cite{whittington1992possible,nauges2006water}).}

The third step maps the share of households using from each source --- connecting alone, sharing with a neighbor, or using from a non-piped source --- into an estimate of each source's relative fixed cost to consumers.  The empirical approach leverages the assumption that given water preferences (from step one) and marginal sharing costs (from step two), any remaining variation in water source choices must be due to differences in their fixed costs.  Since sharing decisions are strategic in nature, I specify a simple game where households sequentially choose their sources and agree to split any surplus from sharing.

This revealed-preference approach to estimating fixed costs is especially relevant in a developing context because on top of the initial connection and monthly service fees, households often face other fixed costs that are difficult to observe, including permitting fees, plumbing maintenance, as well as land-tenure issues in qualifying for a water connection.  In fact, I estimate that the connection and service fees account for only around half of the total fixed costs.  This structural approach complements previous experimental evidence of fixed costs by \cite{devoto2012happiness}.  By considering both source and quantity choices, this model also expands on the demand estimation literature for public utilities that has so far focused almost exclusively on monthly quantity choices (\cite{diakite2009proposal,mcrae2014infrastructure,olmstead2009reduced,szabo2015value}).

@@

With these estimates, I simulate a series of counterfactual policies.  First, I consider a connection fee subsidy policy consistent with recommendations from the World Bank, NGOs, and other researchers to improve water access for the poor (\cite{komives2005water,jimenez2014factors,mcintosh2003asian}).  I discount the fixed fee by 12\% while increasing the tariff schedule proportionally to ensuring that the water provider exactly covers costs.  This policy actually reduces access because households stop sharing connections and substitute towards non-piped sources.  Next, I compute an optimal tariff schedule to maximize consumer surplus while exactly covering production costs.  Compared to the current tariff, I find that a simple, two-part tariff characterized by a fixed fee and low, flat marginal price can produce gains on the order of \input{tables/tpt_improvement_percent_sharing_group}\unskip\% of consumer surplus.\footnote{With quasi-linear utility, consumer surplus for a given tariff is calculated as the difference between household utility under this tariff and household utility from using the non-piped, vendor option.}  These gains are driven both by pricing closer to marginal cost as well as by increasing sharing, which reduces the number of connections installed and maintained by the utility.

Finally, I compute a tariff that maximizes consumer surplus only for the bottom 50\% of users and is intended to capture goals of ensuring minimum usage and access as well as income redistribution.  With sharing, this tariff is ineffective at reallocating surplus from large to small users; however in a counterfactual without sharing, this tariff achieves substantial redistribution.  These findings suggest that informal sharing networks constrain the extent to which non-linear prices can redistribute surplus.

@@

This paper focuses on one of the providers which serves 1.5 million connections and around 5 million people.  Figure~\ref{figure:tariff} shows the regulated tariff in Manila.  The increasing marginal price is thought to cross-subsidize consumption among low-demand, poorer households through higher bills from large users.  The regulator assigns zero marginal price to the first 10 cubic meters of monthly usage, which is called the ``lifeline'' rate.  This policy is intended to provide a minimum level of consumption for all households.  This tariff may also perform efficient price discrimination if low-usage households are more price-sensitive.
\begin{figure}
\centering
\caption{Water Tariff in Manila}\label{figure:tariff}
\begin{center}
\includegraphics[scale=.7]{tables/tariff_structure.pdf}\\
\footnotesize{1 USD = 50 PhP}
\end{center}
\end{figure}
% detail more about how to get water, with the fixed fee and conne{}ction fee?!
On top of this tariff schedule, the regulator also charges a fixed fee independent of usage totaling 225 PhP/month.  This fee includes both the monthly service charge and the initial connection fee that I have put into monthly terms for the duration of the water connection.\footnote{Appendix~\ref{appendix:connectionfeeamortization} provides more details about how this monthly fee is calculated and amortized.}  The fixed fee allows the provider to recover any operating costs associated with serving the connection.  This form of increasing tariff and fixed fee combination is implemented in over 70\% of developing cities as well as over 50\% of developed cities (\cite{hoque2013state}).

@@

By capturing demand heterogeneity, the costs of sharing, as well as source choices, the structural model provides additional insights into how a range of non-linear pricing policies affect both access and welfare.  In the presence of sharing, the model finds that complex non-linear pricing is largely ineffective in achieving either efficiency or equity; sharing provides a resale market, undermining any goals of price discrimination with non-linear pricing.  The structural model also allows me to forecast how sharing costs affect optimal prices.  I find that optimal prices balance competing goals of pricing at marginal cost and taxing sharing when it is costly.

@@

Since sharing networks make usage-based price discrimination more difficult, cities may benefit from considering a broader range of policies to improve welfare from water use.  For example, cities may be able to lower the frictions associated with these sharing networks by providing technical resources for managing plumbing networks or distributing extra water meters to assist households in accurately splitting the bill.  The empirical results also confirm previous findings that substantial barriers exist for households to connect to the network (\cite{devoto2012happiness}).  Cities may be able to similarly achieve substantial welfare gains by streamlining administrative requirements for new connections and by allowing households to pay connection fees in smaller increments.

>>>>

	In developing cities, high costs of connecting to piped water lead many households to use low-quality, non-piped sources.  In response, cities often set low fixed prices so that more households connect and high marginal prices to cover production costs.  This paper identifies a common alternative source --- sharing a neighbor's connection --- that has large implications for optimal water pricing.  With microdata from 1.5 million water connections in Manila, I structurally estimate water demand in the presence of water sharing and solve the regulator's problem: setting prices to maximize consumer surplus while funding production costs.  I find that low fixed and high marginal prices weaken incentives to share water between neighbors.  In contrast, the optimal pricing policy features high fixed and low marginal prices, increases shared connections, and ensures nearly universal access to piped water compared to current prices in Manila.

@@

These estimates allow me to solve a standard regulator pricing problem: setting prices to cover fixed and marginal costs while maximizing consumer surplus.  I also let the regulator place extra weight on surplus for smaller, poorer users.  In the presence of sharing, the standard approach --- low fixed fees and high marginal prices --- excludes many small-users from piped water by reducing incentives to share connections.  Instead, these households substitute toward non-piped water sources, which have relatively lower marginal prices, zero upfront costs, and varying water quality.\footnote{Common non-piped water sources in Manila include deepwells, tanker trucks, and water refilling stations.  Without fixed prices, land-tenure requirements, or administrative expenses, these sources often have low costs of access.}  The opposite approach --- high fixed fees and low marginal prices --- maximizes both access to piped water and welfare, even for smaller users.  High fixed fees are paid in full by households who connect individually and are split between sharing households.  Informal sharing arrangements enable small-users to access piped water from their neighbors for relatively lower marginal prices than those offered by non-piped sources.  By reducing the total number of connections, sharing arrangements also let the city save on installation and maintenance costs, providing an efficient means for extending piped water to low-demand households.\footnote{In this regulatory arrangement, cost savings are passed on to consumers through lower prices.}

@@

To estimate the marginal hassle costs of sharing water, the second step analyzes a quasi-experiment where breaks in service pipes force some households to switch from their own water connections to their neighbors' connections.  I then measure how much less water households use from their neighbors' connections and attribute this gap to the marginal hassle costs of sharing per cubic meter of usage.  This exercise recovers sizable marginal hassle costs per cubic meter on the order of \input{tables/hassle_percentage}\unskip\% of the average marginal water price.\footnote{This revealed-preference approach builds on a descriptive literature that identifies potential sources of water sharing frictions (\cite{whittington1992possible,nauges2006water}).}

The third step maps the share of households using from each source --- connecting alone, sharing with a neighbor, or using from a non-piped source --- into an estimate of each source's relative fixed cost to consumers.  The empirical approach leverages the assumption that given water preferences (from step one) and marginal sharing costs (from step two), any remaining variation in water source choices must be due to differences in their fixed costs.  Since sharing decisions are strategic in nature, I specify a simple game where households sequentially choose their sources and agree to split any surplus from sharing.  By considering both source and quantity choices, this model expands on the demand estimation literature for public utilities that has so far focused almost exclusively on monthly quantity choices (\cite{diakite2009proposal,mcrae2014infrastructure,olmstead2009reduced,szabo2015value}).

This revealed-preference approach to estimating fixed costs is especially relevant in a developing context because on top of the initial connection and monthly service fees, households often face other fixed costs that are difficult to observe, including permitting fees, plumbing maintenance, as well as land-tenure issues in qualifying for a water connection.  In fact, I estimate that the fixed fees paid to the provider (connection and service fees) account for only around half of the total fixed costs.  This structural approach complements previous experimental evidence of fixed costs by \cite{devoto2012happiness}.

@@

With these estimates, I simulate a series of counterfactual policies.  First, I consider a connection fee subsidy policy consistent with recommendations from the World Bank, NGOs, and other researchers to improve water access for the poor (\cite{komives2005water,jimenez2014factors,mcintosh2003asian}).  I discount the fixed fee by 12\% while increasing the tariff schedule proportionally to ensure that the water provider exactly covers costs.  This policy actually reduces access because households stop sharing connections and substitute towards non-piped sources.  Next, I compute an optimal tariff schedule to maximize consumer surplus while exactly covering production costs.  Compared to the current tariff, I find that a simple, two-part tariff characterized by a fixed fee and low, flat marginal price can produce gains on the order of \input{tables/tpt_improvement_percent_sharing_group}\unskip\% of consumer surplus.\footnote{With quasi-linear utility, consumer surplus for a given tariff is calculated as the difference between household utility under this tariff and household utility from using the non-piped, vendor option.}  These gains are driven both by pricing closer to marginal cost as well as by increasing sharing, which reduces the number of connections installed and maintained by the utility.

Finally, I compute a tariff that maximizes consumer surplus only for the bottom 50\% of users and is intended to capture goals of ensuring minimum usage and access as well as income redistribution.  With sharing, this tariff is ineffective at reallocating surplus from large to small users; however in a counterfactual without sharing, this tariff achieves substantial redistribution.  These findings suggest that informal sharing networks may constrain the extent to which non-linear prices can redistribute surplus.

@@

This paper focuses on one of the providers which serves 1.5 million connections and around 5 million people.  Figure~\ref{figure:tariff} shows the regulated tariff in Manila.  The increasing marginal price is thought to cross-subsidize consumption among low-demand, poorer households through higher bills from large users.  The regulator assigns zero marginal price to the first 10 cubic meters of monthly usage, which is called the ``lifeline'' rate.  This policy is intended to provide a minimum level of consumption for all households.  This tariff may also perform efficient price discrimination if low-usage households are more price-sensitive.
\begin{figure}
\centering
\caption{Marginal Prices in Manila}\label{figure:tariff}
\begin{center}
\includegraphics[scale=.7]{tables/tariff_structure.pdf}\\
\footnotesize{1 USD = 50 PhP}
\end{center}
\end{figure}
% detail more about how to get water, with the fixed fee and conne{}ction fee?!
On top of this tariff schedule, the regulator also charges a fixed fee independent of usage totaling 225 PhP/month.  This fee includes both the monthly service charge and the initial connection fee that I have put into monthly terms for the duration of the water connection.\footnote{Appendix~\ref{appendix:connectionfeeamortization} provides more details about how this monthly fee is calculated and amortized.}  The fixed fee allows the provider to recover any operating costs associated with serving the connection.  This form of increasing tariff and fixed fee combination is implemented in over 70\% of developing cities as well as over 50\% of developed cities (\cite{hoque2013state}).

@@

By capturing demand heterogeneity, the costs of sharing, as well as source choices, the structural model provides additional insights into how a range of non-linear pricing policies affect both access and welfare.  In the presence of sharing, the model finds that complex non-linear pricing is largely ineffective at achieving efficiency and equity gains because sharing provides a water resale market, undermining any usage-based price discrimination.  The structural model also allows me to forecast how sharing costs affect optimal prices.  I find that optimal prices balance competing goals of pricing at marginal cost and taxing sharing when it is too costly.

@@

Since sharing networks make it more difficult to implement usage-based price discrimination, cities may benefit from considering a broader range of policies to improve welfare from piped water use.  For example, cities may be able to lower the frictions associated with these sharing networks by providing technical resources for managing plumbing networks or distributing extra water meters to assist households in accurately splitting the bill.  The empirical results also confirm previous findings that substantial barriers exist for households to connect to the network (\cite{devoto2012happiness}).  Cities may be able to similarly achieve substantial welfare gains by streamlining administrative requirements for new connections and by allowing households to pay connection fees in smaller increments.
